LabMate: supporting types for MATLAB

### 2024-03-21 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] rough draft Section 3

%%%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[colorlinks=true]{hyperref}
\usepackage{microtype}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

% matrix types, and their motivation

@@

\section{Supporting dimensional consistency}

@@

\bibliographystyle{ws-procs9x6}
%\bibliography{labmate}

>>>>

%%%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[colorlinks=true]{hyperref}
\usepackage{microtype}
\usepackage{proof}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\newcommand{\istype}[1]{#1\ \textsf{type}}
\newcommand{\isadd}[1]{#1\ \textsf{numeric}}
\newcommand{\ismult}[3]{(#1, #2, #3)\ \textsf{multipliable}}

\newcommand{\append}{+\!\!\!+}
\newcommand{\hjux}[2]{{#1}|{#2}}
\newcommand{\vjux}[2]{\frac{#1}{#2}}

\newcommand{\One}{\mathbf{1}}
\newcommand{\Matrix}[5]{\mathsf{Matrix}\,#1\,#2\,#3\,#4\,#5}
\newcommand{\List}[1]{\mathsf{List}\,#1}

@@

Under the hood, LabMate translates MATLAB expressions to typed
expressions in its own core type theory.
%
Since matrices feature heavily in MATLAB code, the type of matrices
play a central role in this type theory.
%
We make this type rather precise, by allowing the type of each entry
in the matrix to vary according to the row and column of the entry:
%
we parameterise the type of matrices by a type $R$ associated with
each row, a type $C$ associated with each column, and a type $E(x,y)$
which might vary depending on the values of parameters $x : R$ and
$y : C$, as well as two lists $rs : \List{R}$ and $\List{C}$ --- the
idea being that the $i$th entry in the list $rs$ and the $j$th entry
in the list $cs$ determines the type of the $(i, j)$ coordinate in the
matrix.
%
Summarised as an inference rule, matrix types are thus formed as follows:
%
\[
  \infer{\istype{\Matrix{R}{C}{E}{rs}{cs}}}
  {
    \istype{R}
    &\quad
    \istype{C}
    &\quad
    x : R, y : C \vdash \istype{E(x,y)}
    &\quad
    rs : \List{R}
    &\quad
    cs : \List{C}
  }
\]

A notable special case is when both $R$ and $C$ are the unit type
$R = C = \One$, that is, the type with exactly one element.
%
In this case, there is only one type of entries $E$ (because both $x$
and $y$ must be the unique element of the unit type), and the only
information available in the lists $rs : \List{\One}$ and
$cs : \List{\One}$ is their length.
%
Thus, we have recovered $\Matrix{\One}{\One}{E}{m}{n}$ as the type of
$m \times n$ matrices with entries of type $E$, where $n$ and $m$ are
the unique lists over the unit type of length $n$ and $m$
respectively.

Another noteworthy special case which we want to support takes $R$ and
$R$ to be types of physical dimensions, with $E(d_1, d_2)$ to be the
type of physical quantities of dimension $d_1 \cdot d_2^{-1}$.
%
This way, we can reduce dimensional consistency checking \`a la
Kennedy~\cite{kennedyUOM} also for programs involving matrices,
following the work of Hart~\cite{hart}.
%
We will see in \sref{sec:example-revisited} how this can be helpful.

Terms of matrix type in the core type theory are built up from
$1 \times 1$ matrices and pasting matrices horizontally and vertically.
%
Singleton matrices have the following typing rule
\[
  \infer{[e] : \Matrix{R}{C}{E}{[r]}{[c]}}
    {r : R
      &\quad
      c : C
      &\quad
      e : E(r, c)
    }
\]
%
while horizontal and vertical pasting horizontally have the following
typing rules:
\[
  \infer{\hjux{A}{B} : \Matrix{R}{C}{E}{rs}{(cs_A \append cs_B)}}
  {
    A : \Matrix{R}{C}{E}{rs}{cs_A}
    &\quad
    B : \Matrix{R}{C}{E}{rs}{cs_B}
  }
\]
\[
  \infer{\vjux{A}{B} : \Matrix{R}{C}{E}{(rs_A \append rs_B)}{cs}}
  {
    A : \Matrix{R}{C}{E}{rs_A}{cs}
    &\quad
    B : \Matrix{R}{C}{E}{rs_B}{cs}
  }
\]
When comparing matrices for equality, we reassociate horizontal and
vertical pastings, so that $\hjux{\vjux{A}{C}}{\vjux{B}{D}}$ and
$\vjux{\hjux{A}{B}}{\hjux{C}{D}}$ indeed are considered equal.

Two matrices $A$ and $B$ can be added if all entry types $E(x, y)$ are
numeric (for example integers, or floating point numbers), and if $A$
and $B$ have the same matrix type:
%
\[
  \infer{A + B : \Matrix{R}{C}{E}{rs}{cs}}
  {
    x : R, y : C \vdash \isadd{E(x, y)}
    &\quad
    A : \Matrix{R}{C}{E}{rs}{cs}
    &\quad
    B : \Matrix{R}{C}{E}{rs}{cs}
  }
\]
%
Matrix multiplication is trickier, because it does not demand that $A$
and $B$ have the same type --- rather, it demands that $A$ and $B$
have \remph{compatible} types:
%
\[
  \infer{A * B : \Matrix{R}{C}{E''}{rs}{cs}}
  {
    \ismult{E}{E'}{E''}
    &\quad
    A : \Matrix{R}{C}{E}{rs}{ms}
    &\quad
    B : \Matrix{R}{C}{E'}{ms}{cs}
  }
\]
The entry types $E$, $E'$, $E''$ are multipliable if for all $x$, $y$
and $z$, we have that $E(x, y)$ and $E'(y, z)$ supports a
multiplication operation with codomain $E''(x, z)$.
%
This is certainly the case if all entry types involved are simply
integers or floating point numbers, but this constraint can also
ensure that the physical units involved line up.

@@

\section{Supporting dimensional consistency}
\label{sec:example-revisited}

@@

\bibliographystyle{ws-procs9x6}
\bibliography{labmate}
